\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.Rmd
add caret to package comparisons in draft JSS paper

We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024] (and which is by a wide margin the most downloaded from the RStudio CRAN mirror of the CV packages mentioned in the preceding paragraph). Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^k-fold-used-in-rsample]
```{r rsample-example, cache=TRUE}
library(rsample)

@@

library(purrr, warn.conflicts=FALSE)
# apply CV:
rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
# summarize results:
rs_obj$accuracy <- map_dbl(rs_obj$results,
                           function(x) mean(x$correct))
# accuracy is the complement of the Bayes-rule CV criterion:
1 - summary(rs_obj$accuracy) 
```
[^k-fold-used-in-rsample]: The example in the \pkg{rsample} vignette uses 10-fold rather than LOO CV, but otherwise the code from the vignette is only trivially modified here.

@@

@@

Let's compare the computational efficiency of the three implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:
```{r compare-rsample-cv, cache=TRUE}
all.equal(cv.attrition$`CV crit`,
          cv(mod.attrition, k="loo", criterion=BayesRule,
             method="Woodbury")$`CV crit`,
          )
all.equal(cv.attrition$`CV crit`,
          cv(mod.attrition, k="loo", criterion=BayesRule,
             method="hatvalues")$`CV crit`
          )

microbenchmark::microbenchmark(
  rsample={
    rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
    rs_obj$accuracy <- map_dbl(rs_obj$results,
                               function(x) mean(x$correct))
  },
  direct={
    n <- nrow(attrition)
    yhat <- numeric(n)
    for (i in 1:n){
      m <- update(mod.attrition, data=attrition[-i, ])
      yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
    }
    mean(((attrition$Attrition == "Yes") - round(yhat))^2)
  },
  cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
  cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
             method="Woodbury"),
  cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
                  method="hatvalues"),
  times=10
)
```
Computing time for \pkg{rsample} for this problem is similar to direct computation of LOO CV and to \code{cv()} with \code{method="exact"}, but more than an order of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.

>>>>

<!-- We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024] (and which is by a wide margin the most downloaded from the RStudio CRAN mirror of the CV packages mentioned in the preceding paragraph).  -->

We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024], comparing LOO CV computed using the \code{cv()} function in the \pkg{cv} package to LOO CV computed using the \pkg{caret} and \pkg{rsample} packages. At more than 15 million cumulative downloads, \pkg{caret} is by a very wide margin the most downloaded from the RStudio CRAN mirror of the \proglang{R} packages mentioned above. Among the packages that focus more on cross-validation, \pkg{rsample} is the most downloaded, at nearly 4 million downloads.

Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^k-fold-used-in-rsample]
```{r rsample-example, cache=TRUE}
library("rsample")

@@

library("purrr", warn.conflicts=FALSE)
# apply CV:
rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
# summarize results:
rs_obj$accuracy <- map_dbl(rs_obj$results,
                           function(x) mean(x$correct))
# accuracy is the complement of the Bayes-rule CV criterion:
1 - summary(rs_obj$accuracy) 
```
[^k-fold-used-in-rsample]: The example in the \pkg{rsample} vignette uses 10-fold rather than LOO CV, but otherwise the code from the vignette is only trivially modified here.

@@

The \pkg{caret} package doesn't directly support generalized linear models fit by \code{glm()}, but it can be made to do so by defining a custom "method" for the binary logistic regression fit to the \code{attrition} data:
```{r rsample-example-using-caret, cache=TRUE}
library("caret", quietly=TRUE, warn.conflicts=FALSE)

glm_mod <- getModelInfo("glm", regex = FALSE)[[1]]
glm_mod$fit <- function(x, y, wts, param, lev, last, classProbs, ...) {
  x$Attrition <- y
  glm(Attrition ~ JobSatisfaction + Gender + MonthlyIncome,
      family = binomial, data = x)
}

mod <- train(x = attrition[, c("JobSatisfaction", "Gender",
                               "MonthlyIncome")],
             y = attrition$Attrition,
             method = glm_mod,
             trControl = trainControl(method="LOOCV"))
mod # reports complement of Bayes-rule criterion
```

@@

Let's compare the computational efficiency of the various implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:
```{r compare-rsample-cv, cache=TRUE}
all.equal(cv.attrition$`CV crit`,
          cv(mod.attrition, k="loo", criterion=BayesRule,
             method="Woodbury")$`CV crit`,
          )
all.equal(cv.attrition$`CV crit`,
          cv(mod.attrition, k="loo", criterion=BayesRule,
             method="hatvalues")$`CV crit`
          )

microbenchmark::microbenchmark(
  rsample={
    rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
    rs_obj$accuracy <- map_dbl(rs_obj$results,
                               function(x) mean(x$correct))
  },
  caret=train(x = attrition[, c("JobSatisfaction", "Gender",
                                "MonthlyIncome")],
              y = attrition$Attrition,
              method = glm_mod,
              trControl = trainControl(method="LOOCV")),
  direct={
    n <- nrow(attrition)
    yhat <- numeric(n)
    for (i in 1:n){
      m <- update(mod.attrition, data=attrition[-i, ])
      yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
    }
    mean(((attrition$Attrition == "Yes") - round(yhat))^2)
  },
  cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
  cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
             method="Woodbury"),
  cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
                  method="hatvalues"),
  times=10
)
```
Computing time for \pkg{rsample} and \pkg{caret} for this problem is similar to direct computation of LOO CV and to \code{cv()} with \code{method="exact"}, but more than an order of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.

### 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.tex
add caret to package comparisons in draft JSS paper

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 716237
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.28027
bias-adjusted cross-validation criterion = 19.2646
95% CI for bias-adjusted CV criterion = (15.81196, 22.71725)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function(y, yhat) {
  if (!all(y %in% c(0, 1)))
    stop("response values not all 0 or 1")
  if (any(yhat < 0) ||
      any(yhat > 1))
    stop("fitted values outside of interval [0, 1]")
  yhat <- round(yhat)
  result <- mean(y != yhat) # proportion in error
  attr(result, "casewise loss") <- "y != round(yhat)"
  result
}
<bytecode: 0x126d8b808>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

@@

We illustrate with an example adapted from a vignette in the
\pkg{rsample} package, which uses the \code{attrition} data set from the
\pkg{modeldata} package \citep{Kuhn:2024} (and which is by a wide margin
the most downloaded from the RStudio CRAN mirror of the CV packages
mentioned in the preceding paragraph). Here, with minimal explanation,
is how one would perform LOO CV for this example using
\pkg{rsample}:\footnote{The example in the \pkg{rsample} vignette uses
  10-fold rather than LOO CV, but otherwise the code from the vignette
  is only trivially modified here.}

\begin{CodeChunk}
\begin{CodeInput}
R> library(rsample)
R> 
R> data("attrition", package = "modeldata")
R> nrow(attrition)
\end{CodeInput}
\begin{CodeOutput}
[1] 1470
\end{CodeOutput}
\begin{CodeInput}
R> # formula for model to be fit; Attrition is binary:
R> mod_form <- Attrition ~ JobSatisfaction + Gender + MonthlyIncome
R> 
R> rs_obj <- loo_cv(attrition) # original uses vfold_cv()
R> 
R> # user-supplied function:
R> holdout_results <- function(splits, ...) {
+   mod <- glm(..., data=analysis(splits), family=binomial)
+   holdout <- assessment(splits)
+   res <- broom::augment(mod, newdata=holdout)
+   lvls <- levels(holdout$Attrition)
+   predictions <- factor(ifelse(res$.fitted > 0, lvls[2], lvls[1]),
+                         levels=lvls)
+   res$correct <- predictions == holdout$Attrition
+   res
+ }
R> 
R> library(purrr, warn.conflicts=FALSE)
R> # apply CV:
R> rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
R> # summarize results:
R> rs_obj$accuracy <- map_dbl(rs_obj$results,
+                            function(x) mean(x$correct))
R> # accuracy is the complement of the Bayes-rule CV criterion:
R> 1 - summary(rs_obj$accuracy) 
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 1.0000  0.0000  0.0000  0.1612  0.0000  0.0000 
\end{CodeOutput}
\end{CodeChunk}

@@

@@

Let's compare the computational efficiency of the three implementations,
also showing the use of \code{method="Woodbury"} and
\code{method="hatvalues"} for \code{cv()}, which, in this example,
produce numerically identical results to the default
\code{method="exact"}:

\begin{CodeChunk}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury")$`CV crit`,
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="hatvalues")$`CV crit`
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   rsample={
+     rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
+     rs_obj$accuracy <- map_dbl(rs_obj$results,
+                                function(x) mean(x$correct))
+   },
+   direct={
+     n <- nrow(attrition)
+     yhat <- numeric(n)
+     for (i in 1:n){
+       m <- update(mod.attrition, data=attrition[-i, ])
+       yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
+     }
+     mean(((attrition$Attrition == "Yes") - round(yhat))^2)
+   },
+   cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
+   cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury"),
+   cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
+                   method="hatvalues"),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(rsample = {: less accurate nanosecond
times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: milliseconds
         expr         min          lq        mean      median         uq
      rsample 5707.408772 5711.011032 5810.464433 5828.739437 5853.64979
       direct 5187.215696 5206.152940 5281.503347 5230.005408 5348.24496
     cv.exact 4659.560948 4691.022749 4735.789317 4737.251540 4790.76759
      cv.wood   99.232628  104.541882  106.917914  106.051051  108.49436
 cv.hatvalues    1.187401    1.224875    2.751965    1.356813    1.56374
        max neval
 5999.09487    10
 5500.60457    10
 4811.15176    10
  119.72115    10
   15.35548    10
\end{CodeOutput}
\end{CodeChunk}

Computing time for \pkg{rsample} for this problem is similar to direct
computation of LOO CV and to \code{cv()} with \code{method="exact"}, but
more than an order of magnitude slower than \code{cv()} with
\code{method="Woodbury"}, and more than three orders of magnitude slower
then \code{cv()} with \code{method="hatvalues"}.

>>>>

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 799417
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.20609
bias-adjusted cross-validation criterion = 19.19442
95% CI for bias-adjusted CV criterion = (15.73743, 22.6514)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function(y, yhat) {
  if (!all(y %in% c(0, 1)))
    stop("response values not all 0 or 1")
  if (any(yhat < 0) ||
      any(yhat > 1))
    stop("fitted values outside of interval [0, 1]")
  yhat <- round(yhat)
  result <- mean(y != yhat) # proportion in error
  attr(result, "casewise loss") <- "y != round(yhat)"
  result
}
<bytecode: 0x1118adc78>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

@@

We illustrate with an example adapted from a vignette in the
\pkg{rsample} package, which uses the \code{attrition} data set from the
\pkg{modeldata} package \citep{Kuhn:2024}, comparing LOO CV computed
using the \code{cv()} function in the \pkg{cv} package to LOO CV
computed using the \pkg{caret} and \pkg{rsample} packages. At more than
15 million cumulative downloads, \pkg{caret} is by a very wide margin
the most downloaded from the RStudio CRAN mirror of the \proglang{R}
packages mentioned above. Among the packages that focus more on
cross-validation, \pkg{rsample} is the most downloaded, at nearly 4
million downloads.

Here, with minimal explanation, is how one would perform LOO CV for this
example using \pkg{rsample}:\footnote{The example in the \pkg{rsample}
  vignette uses 10-fold rather than LOO CV, but otherwise the code from
  the vignette is only trivially modified here.}

\begin{CodeChunk}
\begin{CodeInput}
R> library("rsample")
R> 
R> data("attrition", package = "modeldata")
R> nrow(attrition)
\end{CodeInput}
\begin{CodeOutput}
[1] 1470
\end{CodeOutput}
\begin{CodeInput}
R> # formula for model to be fit; Attrition is binary:
R> mod_form <- Attrition ~ JobSatisfaction + Gender + MonthlyIncome
R> 
R> rs_obj <- loo_cv(attrition) # original uses vfold_cv()
R> 
R> # user-supplied function:
R> holdout_results <- function(splits, ...) {
+   mod <- glm(..., data=analysis(splits), family=binomial)
+   holdout <- assessment(splits)
+   res <- broom::augment(mod, newdata=holdout)
+   lvls <- levels(holdout$Attrition)
+   predictions <- factor(ifelse(res$.fitted > 0, lvls[2], lvls[1]),
+                         levels=lvls)
+   res$correct <- predictions == holdout$Attrition
+   res
+ }
R> 
R> library("purrr", warn.conflicts=FALSE)
R> # apply CV:
R> rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
R> # summarize results:
R> rs_obj$accuracy <- map_dbl(rs_obj$results,
+                            function(x) mean(x$correct))
R> # accuracy is the complement of the Bayes-rule CV criterion:
R> 1 - summary(rs_obj$accuracy) 
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 1.0000  0.0000  0.0000  0.1612  0.0000  0.0000 
\end{CodeOutput}
\end{CodeChunk}

@@

The \pkg{caret} package doesn't directly support generalized linear
models fit by \code{glm()}, but it can be made to do so by defining a
custom ``method'' for the binary logistic regression fit to the
\code{attrition} data:

\begin{CodeChunk}
\begin{CodeInput}
R> library("caret", quietly=TRUE, warn.conflicts=FALSE)
R> 
R> glm_mod <- getModelInfo("glm", regex = FALSE)[[1]]
R> glm_mod$fit <- function(x, y, wts, param, lev, last, classProbs, ...) {
+   x$Attrition <- y
+   glm(Attrition ~ JobSatisfaction + Gender + MonthlyIncome,
+       family = binomial, data = x)
+ }
R> 
R> mod <- train(x = attrition[, c("JobSatisfaction", "Gender",
+                                "MonthlyIncome")],
+              y = attrition$Attrition,
+              method = glm_mod,
+              trControl = trainControl(method="LOOCV"))
R> mod # reports complement of Bayes-rule criterion
\end{CodeInput}
\begin{CodeOutput}
Generalized Linear Model

1470 samples
   3 predictor
   2 classes: 'No', 'Yes'

No pre-processing
Resampling: Leave-One-Out Cross-Validation 
Summary of sample sizes: 1469, 1469, 1469, 1469, 1469, 1469, ... 
Resampling results:

  Accuracy   Kappa
  0.8387755  0    
\end{CodeOutput}
\end{CodeChunk}

@@

Let's compare the computational efficiency of the various
implementations, also showing the use of \code{method="Woodbury"} and
\code{method="hatvalues"} for \code{cv()}, which, in this example,
produce numerically identical results to the default
\code{method="exact"}:

\begin{CodeChunk}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury")$`CV crit`,
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="hatvalues")$`CV crit`
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   rsample={
+     rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
+     rs_obj$accuracy <- map_dbl(rs_obj$results,
+                                function(x) mean(x$correct))
+   },
+   caret=train(x = attrition[, c("JobSatisfaction", "Gender",
+                                 "MonthlyIncome")],
+               y = attrition$Attrition,
+               method = glm_mod,
+               trControl = trainControl(method="LOOCV")),
+   direct={
+     n <- nrow(attrition)
+     yhat <- numeric(n)
+     for (i in 1:n){
+       m <- update(mod.attrition, data=attrition[-i, ])
+       yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
+     }
+     mean(((attrition$Attrition == "Yes") - round(yhat))^2)
+   },
+   cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
+   cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury"),
+   cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
+                   method="hatvalues"),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(rsample = {: less accurate nanosecond
times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: milliseconds
         expr         min          lq        mean      median          uq
      rsample 5853.035606 5948.291603 6163.101903 6121.042951 6144.223100
        caret 6804.763276 6951.219335 7097.717763 7061.439656 7255.566640
       direct 5310.127259 5350.848336 5554.513023 5484.144809 5690.480487
     cv.exact 4751.498733 4797.980556 4879.481352 4870.303039 4927.941413
      cv.wood  101.606159  103.891294  108.285899  108.577430  112.075263
 cv.hatvalues    1.206999    1.471613    1.514548    1.554679    1.614416
         max neval
 6981.899881    10
 7541.571242    10
 6081.412412    10
 5118.834994    10
  116.574603    10
    1.759433    10
\end{CodeOutput}
\end{CodeChunk}

Computing time for \pkg{rsample} and \pkg{caret} for this problem is
similar to direct computation of LOO CV and to \code{cv()} with
\code{method="exact"}, but more than an order of magnitude slower than
\code{cv()} with \code{method="Woodbury"}, and more than three orders of
magnitude slower then \code{cv()} with \code{method="hatvalues"}.
